---
execute: 
  echo: true
  eval: false
  warning: false
---

# Concatenation-Based Phylogeny Inference

In this chapter, we outline the process of building a phylogenetic tree using the concatenation-based approach. This technique consists of combining sequences from multiple genes into a single dataset, enabling a comprehensive reconstruction of evolutionary relationships. The resulting tree will be used as a point of comparison to assess the accuracy and robustness of the new phylogenetic method developed in this project.

```{r here}
#| message: false
#| eval: true

library(here)
library(cogeqc)
library(Biostrings)
library(tidyverse)
library(ape)
library(msa)
library(Herper)
library(ggtree)

set.seed(123) # for reproducibility
options(timeout = 1e8)

source(here("code", "utils.R")) # helper functions
```

## The multiple sequence alignment (MSA)

We'll first load the BUSCO genes. These are conserved genes that are expected to be present as single-copies in the genomes. There are 362 BUSCO genes that we will be using.

```{r}
busco_genes <- readRDS(here("data","busco_seqs.rds"))
```

Each BUSCO gene was aligned with MUSCLE, converted to amino acid format, and trimmed to remove poorly aligned regions (>50% gaps).

```{r}
# Perform MSA for each BUSCO gene with MUSCLE
msas <- lapply(busco_genes, msa, method = "Muscle")

# Convert MSAs to `AAStringSet` class
msas_aa <- lapply(msas, as, "AAStringSet")

# Trim MSAs with remove columns with >50% of gaps
trimmed_msas <- lapply(msas_aa, trim_alignment, max_gap = 0.5)
```

## Concatenation

IQ-TREE 2 was installed and added to the system PATH.

```{r}
Sys.setenv(
  PATH = paste(
    Sys.getenv("PATH"),
    "C:\\Users\\emile\\AppData\\Local\\Microsoft\\WindowsApps\\iqtree-2.4.0-Windows\\bin",
    sep = ";"
  )
)
```

The trimmed alignments were concatenated into a single supermatrix, which was then used to infer a species tree with a defined outgroup using a multi-threaded approach.

```{r}
# Concatenate alignments
conc_msa <- Reduce(xscat, trimmed_msas)
names(conc_msa) <- names(trimmed_msas[[1]])

# Take a look at the concatenated alignment
head(conc_msa)

# Infer species tree from concatenated alignment
outgroup <- "Rhizobium_rhizophilum_gca_004912145.ASM491214v1"
tree_conc <- infer_species_tree(
    list(conc = conc_msa),
    outgroup, 
    threads = 4
)
```

## Tree visualization

The inferred phylogenetic tree was exported in Newick format for downstream analysis and reproducibility.

```{r}
# Save the tree file
ape::write.tree(tree_conc, file = here("products", "result_files", "tree_conca.nwk"))
```

The saved tree was reloaded as a phylogenetic object and visualized to inspect topology and labeling, with improved tip readability.

```{r}
# Read tree as a `phylo` object
tree_conc <- read.tree(here("products", "result_files", "tree_conca.nwk"))

# Plot tree
plot(tree_conc, show.node.label = TRUE)

# Tree less overlap
p_conc <- ggtree(tree_conc) +
  geom_tiplab(fontface="italic")
```

## Session info {.unnumbered}

This document was created under the following conditions:

```{r sessioninfo}
#| eval: true
#| echo: false
sessioninfo::session_info()
```

## References {.unnumbered}
