---
execute: 
  echo: true
  eval: false
  warning: false
---

# Coalescence-Based Phylogony Inference

In this chapter, we infer a phylogenetic tree using a coalescence‑based approach, which reconstructs evolutionary relationships by integrating information from multiple individual gene trees. Rather than merging all gene sequences into a single alignment, this method treats each gene independently and explicitly accounts for variation in their evolutionary histories. This makes coalescent‑based inference particularly well suited for handling heterogeneous signals across loci.


```{r here}
#| message: false
#| eval: true


library(here)
library(cogeqc)
library(Biostrings)
library(tidyverse)
library(ape)
library(msa)
library(BiocManager)
library(S4Vectors)
library(IRanges)
library(GenomeInfoDb)
library(GenomicRanges)
library(ggplot2)
library(Herper)
library(ASTRAL)
library(treeio)
library(ggtree)
annotation <- readRDS(here("data" ,"annotation.rds"))
annotation

busco_genes <- readRDS(here("data","busco_seqs.rds"))
busco_genes

proteomes <- readRDS(here ("data", "proteomes.rds"))
proteomes

species_metadata <- readRDS(here ("data", "species_metadata"))
species_metadata


set.seed(123) # for reproducibility
options(timeout = 1e8)

source(here("utils.R")) # helper functions
Sys.setenv(
  PATH = paste(
    Sys.getenv("PATH"), "C:\\Users\\woutd\\AppData\\Local\\Microsoft\\WindowsApps\\iqtree-2.4.0-Windows\\bin" , sep = ";"
  )
)


```

## Performing an msa

After all the data was loaded, a multiple seqeunce allignment can be created by using the Muscle method. To use the msa later it is also easier to convert the output into an AAStringSet
```{r here}
# Perform MSA for each BUSCO gene with MUSCLE
msas <- lapply(busco_genes, msa, method = "Muscle")

# Convert MSAs to `AAStringSet` class
msas_aa <- lapply(msas, as, "AAStringSet")

# Take a look at the MSAs for the first 3 genes
head(msas_aa, 3)
```

## Trimming the msa tree, so all the gaps are removed

As the multiple sequence alignment still contains multiple gaps which that can interfere when making a tree. A function called "trimmed_msas" was used. When inferring the tree, a problem arose where some genes wouldn't infer with some species, to solve this, these genes were removed. The removed genes can ben found in the vector called "bad"   
```{r}
# Trim MSAs with remove columns with >50% of gaps
trimmed_msas <- lapply(msas_aa, trim_alignment, max_gap = 0.5)

# Remove bad genes
bad <- c("102328at356", "125305at356" , "128354at356","129387at356","133912at356","134247at356","140338at356","141504at356","20294at356","46097at356","47774at356","53126at356","68921at356","80701at356", "87562at356", "87990at356","95094at356","103547at356","113565at356","116138at356","119266at356","120363at356","121617at356","")
trimmed_msas_final <- trimmed_msas[!names(trimmed_msas) %in% bad]

# Take a look at trimmed MSAs
head(trimmed_msas, 3)
```

## Inferrring the final tree

After receiving the actual data, a tree can be made. In this chapter the coalescent method is used, which means that for each BUSCO gene available a species tree is made. in total there were 339 genes and consequently there were also 339 different trees
```{r}
# Infer single-locus trees
outgroup <- "Rhizobium_rhizophilum_gca_004912145.ASM491214v1"
tree_sl <- infer_singlelocus_trees(
    trimmed_msas_final,
    outgroup,
    outdir = here("products", "sl_trees"),
    threads = 8
)

# Read tree
tree_sl <- read.tree(here("products", "sl_trees.treefile" ))

tree_sl
```

## Constructing the final tree

As there is no efficient and correct way to compare all of the 339 different trees, a single one was made. This tree was constructed by the ASTRAL program. 
```{r}

system("java -jar astral.5.7.8.jar -i sl.treefile -o astral.tree")

# Read species tree created with ASTRAL
astral_tree <- read.tree(here("output", "sl_trees", "astral.tree"))

complete_tree <- read.tree(here("astral_tree_gcf.tree"))


# Plot tree
plot(complete_tree, show.node.label = TRUE)

```



## Session info {.unnumbered}

This document was created under the following conditions:

```{r sessioninfo}
#| eval: true
#| echo: false
sessioninfo::session_info()
```

## References {.unnumbered}
