---
execute: 
  echo: true
  eval: false
  warning: false
---

# Genomic data acquisition

In this chapter, we will describe the code to obtain genomic data from 
Ensembl Bacteria. In this project, we will use genomes from *Agrobacterium sp.*
species. As of March 27, there are 70 species on Ensembl Bacteria. And as an outgroup we will use the genome from *Rhizobium sp.*.

```{r here}
#| message: false
#| eval: true

library(here)
library(biomartr)
library(tidyverse)

set.seed(123) # for reproducibility
```

## Download data from Ensembl Bacteria

We will download species metadata from Ensembl Bacteria, and then use
species names to download proteomes and annotation using __biomartr__.

```{r}
# Get metadata for *Agrobacterium sp.*
bacteria_metadata <- read_tsv(
    "https://ftp.ensemblgenomes.ebi.ac.uk/pub/bacteria/current/species_EnsemblBacteria.txt",
    show_col_types = FALSE
) |>
    janitor::clean_names() |>
    filter(str_detect(species, "agrobacterium"))

# Download proteomes
proteome_dir <- here("data", "proteomes")
if(!dir.exists(proteome_dir)) { dir.create(proteome_dir, recursive = TRUE) }

getProteomeSet(
    db = "ensembl", organisms = bacteria_metadata$species, path = proteome_dir
)

# Download annotation
annotation_dir <- here("data", "annotation")
if(!dir.exists(annotation_dir)) { dir.create(annotation_dir, recursive = TRUE) }

getGFFSet(
    db = "ensembl", organisms = bacteria_metadata$species, path = annotation_dir,
    skip_bacteria = FALSE
)
```

Now, we will load the proteomes as a list of `AAStringSet` objects,
and annotation as a list of `GRangesList` objects.

```{r}
# Load annotation
annot_files <- list.files(
    here("data", "annotation"), pattern = ".gff3", full.names = TRUE
)
names(annot_files) <- gsub("\\.62\\._ensembl.*", "", basename(annot_files))
annot <- lapply(annot_files, function(x) {
    ranges <- rtracklayer::import(x)
    ranges <- ranges[ranges$type == "gene"]
    
    return(ranges)
})

# Load proteomes
seq_files <- list.files(
    here("data", "proteomes"), pattern = ".fa", full.names = TRUE
)
names(seq_files) <- gsub("\\.pep.*", "", basename(seq_files))
seq <- lapply(seq_files, function(x) {
    s <- Biostrings::readAAStringSet(x)
    names(s) <- gsub(" .*", "", gsub(".* gene:", "", names(s)))
    
    return(s)
})
```

## Wrapping up {.unnumbered}

Finally, we will save objects created in this chapter for further reuse.

```{r}
# R objects
## list of `AAStringSet`: proteomes
saveRDS(
    seq, compress = "xz",
    file = here("data", "proteomes.rds")
)

## list of `GRanges`: annotation
saveRDS(
    annot, compress = "xz",
    file = here("data", "annotation.rds")
)

## Species metadata
saveRDS(
    bacteria_metadata, compress = "xz",
    file = here("data", "species_metadata.rds")
)
```

## Session info {.unnumbered}

This document was created under the following conditions:

```{r sessioninfo}
#| eval: true
#| echo: false
sessioninfo::session_info()
```

## References {.unnumbered}
